---
title: "Assignment #1"
format: html
---
```{r setup, include=FALSE}
# Set up a consistent path for all chunks of code
knitr::opts_knit$set(root.dir = normalizePath("../../"))
```

```{r, message=FALSE, warning=FALSE, results='hide'}
library(data.table)
library(ggplot2)
library(seqminer)
library(HardyWeinberg)
library(dplyr)
```


## Assignment 1
You only need to write lines of code for each question. When answering questions that ask you to identify or interpret something, the length of your response doesn’t matter. For example, if the answer is just ‘yes,’ ‘no,’ or a number, you can just give that answer without adding anything else.

We will go through comparable code and concepts in the live learning session. If you run into trouble, start by using the help help() function in R, to get information about the datasets and function in question. The internet is also a great resource when coding (though note that no outside searches are required by the assignment!). If you do incorporate code from the internet, please cite the source within your code (providing a URL is sufficient).

Please bring questions that you cannot work out on your own to office hours, work periods or share with your peers on Slack. We will work with you through the issue.

You will need to install PLINK and run the analyses. Please follow the OS-specific setup guide in [`SETUP.md`](../../SETUP.md). PLINK is a free, open-source whole genome association analysis toolset, designed to perform a range of basic, large-scale analyses in a computationally efficient manner.

#### Question 1: Data inspection

Before fitting any models, it is essential to understand the data. Use R or bash code to answer the following questions about the `gwa.qc.A1.fam`, `gwa.qc.A1.bim`, and `gwa.qc.A1.bed` files, available at the following Google Drive link: <https://drive.google.com/drive/folders/11meVqGCY5yAyI1fh-fAlMEXQt0VmRGuz?usp=drive_link>. Please download all three files from this link and place them in `02_activities/data/`.

(i) Read the .fam file. How many samples does the dataset contain?
```{r}
fam <- fread("02_activities/data/gwa.qc.A1.fam")
head(fam)
```

```{r}
nrow(fam)
```

There are 4000 samples in the dataset.

(ii) What is the 'variable type' of the response variable (i.e.Continuous or binary)?

``` {r}        
# Assign the standard PLINK column names
colnames(fam) <- c("FID", "IID", "PAT", "MAT", "SEX", "PHENOTYPE")
head(fam)
```

Continuous

(iii) Read the .bim file. How many SNPs does the dataset contain?

```{r}      
# Read the .bim file
bim <- fread("02_activities/data/gwa.qc.A1.bim")

# Assign the standard PLINK column names to the bim file
colnames(bim) <- c("CHR", "SNP", "CM", "BP", "A1", "A2")
head(bim)
```

```{r}
# Count the number of SNPs (rows)
nrow(bim)
```

There are 101083 SNPS in the dataset.


#### Question 2: Allele Frequency Estimation

(i) Load the genotype matrix for SNPs rs1861, rs3813199, rs3128342, and rs11804831 using additive coding. What are the allele frequencies (AFs) for these four SNPs?

```{bash,results='hide',message=FALSE, warning=FALSE}
# Create a text file with four target SNPs
printf "%s\n" rs1861 rs3813199 rs3128342 rs11804831 > 02_activities/data/q2_snplist.txt

# Use PLINK to extract them and create an additive coded .raw file
plink2 --bfile 02_activities/data/gwa.qc.A1 \
      --extract 02_activities/data/q2_snplist.txt \
      --recode A \
      --out 02_activities/data/q2_geno_additive
```

```{r}
# Load the additive genotype matrix 
geno_q2 <- fread("02_activities/data/q2_geno_additive.raw")

# Calculate Allele Frequencies (AFs)
allele_freq <- apply(geno_q2[, 7:ncol(geno_q2)], 2, function(x) {
  sum(x, na.rm = TRUE) / (2 * sum(!is.na(x)))
})

print("Allele Frequencies:")
print(allele_freq)
```

The AFS for the four SNPs are as follows:
 rs3813199_G rs11804831_T  rs3128342_C     rs1861_C 
   0.9430874    0.8456588    0.6948789    0.9460141 

(ii) What are the minor allele frequencies (MAFs) for these four SNPs?

``` {r}        
# Calculate Minor Allele Frequencies (MAFs)
# The MAF = 1 - AF).
maf <- ifelse(allele_freq > 0.5, 1 - allele_freq, allele_freq)

print("Minor Allele Frequencies:")
print(maf)
```

The MAFs for the four SNPs are as follows:
 rs3813199_G rs11804831_T  rs3128342_C     rs1861_C 
  0.05691262   0.15434124   0.30512109   0.05398587 

#### Question 3: Hardy–Weinberg Equilibrium (HWE) Test

(i) Conduct the Hardy–Weinberg Equilibrium (HWE) test for all SNPs in the .bim file. Then, load the file containing the HWE p-value results and display the first few rows of the resulting data frame.

``` {bash,results='hide',message=FALSE, warning=FALSE}
plink2 --bfile ./02_activities/data/gwa.qc.A1 --hardy --out ./02_activities/data/gwa_qc_A1_hwe
```

``` {r}
# Load the HWE results generated by PLINK
hwe_results <- fread("02_activities/data/gwa_qc_A1_hwe.hardy")

# Display the first few rows of the resulting data frame
head(hwe_results)
```

(ii) What are the HWE p-values for SNPs rs1861, rs3813199, rs3128342, and rs11804831?

``` {r}        
# Define the SNPs
target_snps <- c("rs1861", "rs3813199", "rs3128342", "rs11804831")

# Subset the results
hwe_pvalues <- hwe_results[hwe_results$ID %in% target_snps, c("ID", "P")]

print(hwe_pvalues)
```

The p-values are as follows:
rs3813199: 1.000000
rs11804831: 0.113354
rs3128342: 0.330273
rs1861: 0.274719

#### Question 4: Genetic Association Test

(i) Conduct a linear regression to test the association between SNP rs1861 and the phenotype. What is the p-value?

```  {r}
# Conduct linear regression for SNP rs1861
model <- lm(PHENOTYPE ~ rs1861_C, data = geno_q2)
summary (model)
```

The p value is p<0.001 (<2e-16) for the association between SNP rs1861 and the phenotype.

(ii) How would you interpret the beta coefficient from this regression?

The estimate is 0.974, which means that for each additional copy of the 'C' allele, the coninous phenotype is expected to increase by about 0.974 units. 

(iii) Plot the scatterplot of phenotype versus the genotype of SNP rs1861. Add the regression line to the plot.

``` {r}
plot1 <- ggplot(geno_q2, aes(x = rs1861_C, y = PHENOTYPE)) +
  geom_point(alpha = 0.5, color = "blue", position = position_jitter(width = 0.1)) + 
  geom_smooth(method = "lm", color = "red", se = FALSE, size = 1.2) +
  labs(
    title = "Linear Regression: Phenotype vs. rs1861 Genotype (Additive)",
    x = "Genotype (0/1/2)",
    y = "Phenotype"
  ) +
  theme_minimal()

print(plot1)
```

(iv) Convert the genotype coding for rs1861 to recessive coding.

```    {r}     
# Create a copy of the dataset 
geno_q2_rec <- geno_q2

# Convert the rs1861_C column to recessive coding
geno_q2_rec$rs1861_C <- ifelse(geno_q2_rec$rs1861_C == 0, 1, 0)
table(geno_q2_rec$rs1861_C)
```

(v) Conduct a linear regression to test the association between the recessive-coded rs1861 and the phenotype. What is the p-value?

```   {r}      
# Fit the linear regression model for the recessive genotype
model_rec <- lm(PHENOTYPE ~ rs1861_C, data = geno_q2_rec)

summary(model_rec)
```

the p value is p<0.001 (<2e-16) for the association between the recessive-coded rs1861 and the phenotype.

(vi) Plot the scatterplot of phenotype versus the recessive-coded genotype of rs1861. Add the regression line to the plot.

```       {r}  
# Create the scatterplot with the linear regression line for the recessive model
p_rec <- ggplot(geno_q2_rec, aes(x = rs1861_C, y = PHENOTYPE)) +
  geom_point(alpha = 0.5, color = "blue", position = position_jitter(width = 0.1)) + 
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth = 1.2) +
  labs(
    title = "Linear Regression: Phenotype vs. rs1861 Genotype (Recessive)",
    x = "Genotype (0/1)",
    y = "Phenotype"
  ) +
  theme_minimal()

print(p_rec)
```

(vii) Which model fits better? Justify your answer.

```   {r}     
AIC(model, model_rec)
```

          df      AIC
model      3 11276.91
model_rec  3 11291.74

Therefore the additive model has a lower AIC value (11276.91) compared to the recessive model (11291.74), suggesting that the additive model fits the data better than the recessive model.

### Criteria

| Criteria | Complete | Incomplete |
|------------------------|------------------------|------------------------|
| **Data Inspection** | Correct sample/SNP counts and variable type identified. | Missing or incorrect counts or variable type. |
| **Allele Frequency Estimation** | Correct allele and minor allele frequencies computed. | Frequencies missing or wrong. |
| **Hardy–Weinberg Equilibrium Test** | Correct PLINK command and p-value extraction in R. | PLINK command or extraction incorrect/missing. |
| **Genetic Association Test** | Correct regressions, plots, coding, and interpretation. | Regression, plots, or interpretation missing/incomplete. |

### Submission Information

📌 Please review our [Assignment Submission Guide](https://github.com/UofT-DSI/onboarding/blob/main/onboarding_documents/submissions.md) for detailed instructions on how to format, branch, and submit your work. Following these guidelines is crucial for your submissions to be evaluated correctly.

#### Note:

If you like, you may collaborate with others in the cohort. If you choose to do so, please indicate with whom you have worked with in your pull request by tagging their GitHub username. Separate submissions are required.

------------------------------------------------------------------------

#### Submission Parameters

-   Submission Due Date: `11:59 PM – 16/03/2026`

-   Branch name for your repo should be: `assignment-1`

-   What to submit for this assignment:

    -   Populate this Quarto document (`assignment_1.qmd`).
    -   Render the document with Quarto: `quarto render assignment_1.qmd`.
    -   Submit both `assignment_1.qmd` and the rendered HTML file `assignment_1.html` in your pull request.

-   What the pull request link should look like for this assignment: `https://github.com/<your_github_username>/gen_data/pull/<pr_id>`

    -   Open a private window in your browser. Copy and paste the link to your pull request into the address bar. Make sure you can see your pull request properly. This helps the technical facilitator and learning support team review your submission easily.

------------------------------------------------------------------------

Checklist:

-   Created a branch with the correct naming convention.
-   Ensured that the repository is public.
-   Reviewed the PR description guidelines and adhered to them.
-   Verified that the link is accessible in a private browser window.
-   Confirmed that both `assignment_1.qmd` and `assignment_1.html` are included in the pull request.

If you encounter any difficulties or have questions, please don't hesitate to reach out to our team via our Slack help channel. Our technical facilitators and learning support team are here to help you navigate any challenges.
